Generating Random Logic Programs Using Constraint Programming

### 2020-02-13 paper/paper.tex
Editing

How confidently can we claim that an algorithm works well if it is only tested
on a few types of problems? Perhaps the `inferior' algorithm is actually
better in some specific circumstances. Maybe there are identifiable
types of inputs that make every algorithm default to exponential behaviour
that could be easily overcome with the right strategy. At present, most
inference algorithms for probabilistic logic programs are only evaluated on at
most four different problems: sometimes just a single network
\citep{DBLP:journals/tplp/KimmigDRCR11,DBLP:journals/corr/abs-1112-3785,DBLP:conf/iclp/KimmigCRDR08},
sometimes two
\citep{DBLP:journals/corr/abs-1009-3798,DBLP:conf/ecai/BruynoogheMKGVJR10} or
four networks \citep{DBLP:conf/ijcai/VlasselaerBKMR15} coming from a range of
areas such as social networks and citation/genetic/biological data sets. In
order to better understand the strengths and weaknesses of these algorithms,
along with real data they should also be evaluated on a range of synthetic
problems that accurately represent the potential complexities that have to be
handled by a well-designed inference algorithm. Furthermore, random logic
program generators can be useful in combination with methods that generate
random data with probabilistic logic programs \citep{DBLP:conf/soict/Dries15}
and can be used as a component of learning.

Most current approaches to generating random logic programs are restrictive,
e.g., limited to clauses with only two literals
\citep{DBLP:conf/lpnmr/NamasivayamT09}, or to clauses of the form $a \gets \neg
b$ \citep{DBLP:journals/tocl/WenWSL16}, but others are more expressive, e.g.,
defining a program only by the (maximum) number of atoms in the body and the
total number of rules \citep{DBLP:conf/iclp/ZhaoL03}. We introduce a way to
generate random logic programs using a constraint solver such as Choco
\citep{choco}. The same model can generate both probabilistic programs directly
in the syntax of ProbLog \citep{DBLP:conf/ijcai/RaedtKT07} as well as
non-probabilistic Prolog programs. For generated probabilistic programs to be
valid, we use a custom constraint to detect negative cycles (as described in
\cref{sec:cycles}). A major advantage of our constraint-based approach is that
one can easily add additional constraints to the model. To demonstrate that, in
\cref{sec:independence} we present a custom constraint with propagation and
entailment algorithms that can ensure predicate independence.
\cref{sec:counting} also presents a combinatorial argument for correctness,
where we produce combinatorial expressions that count the number of programs
that the model should produce for various parameter values. Finally,
\cref{sec:experiments} shows how the model scales when tasked with producing
more complicated programs and identifies the relationships between parameter
values and the empirical hardness of the program generation task.

% TODO (Vaishak):
Overall, [main] contributions are concerned purely with logic programming-based
languages and frameworks, which capture a major fragment of statistical
relational learning \citep{DBLP:series/synthesis/2016Raedt}.
However, since probabilistic logic programming [languages] are closely related
to other [endeavours] in machine learning, including (imperative) probabilistic
programming (e.g., see discussions in \citep{DBLP:journals/ml/RaedtK15}), our
results [may] provide the [footing] to explore broader questiosn on generating
and testing programs/algorithms in machine learning.

@@

It remains to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate unnecessary symmetries
and adhere to our desired format. First, we can recognise that the order of the
elements in the $\variable{structure}$ array does not matter, i.e., the
structure is only defined by how the elements link to each other, so we can add
a constraint saying that:

@@

% TODO (Vaishak):
% Now we add a theorem that states that the programs generated by the algorithm,
% regardless of whether they involve negative cycles or not, are well-defined?
% That is, in the non-probabilistic case, they are satisfiable, and, in the
% probabilistic case, for both negative cycles & without, they induce programs
% that are satisfiable + well-defined measure on the set of satisfying
% assignments [proofs in the appendix?]

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of thrashing,
i.e., situations where a contradiction is being `fixed' by making changes that
have no hope to fix the contradiction. Thus, we partition all decision variables
into an ordered list of groups, and require the values of all variables from one
group to be determined before moving to the next group. Within each group, we
use the `fail first' variable ordering heuristic. The first group consists of
all head predicates. Afterwards, we handle all remaining decision variables from
the first clause before proceeding to the next. The decision variables within
each clause are divided into:
\begin{enumerate*}
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 20, 40,
80, \dots$ contradictions.

We ran close to \num{400000} experiments, investigating whether the model is
efficient enough to generate reasonably-sized programs and gaining insight into
what parameter values make the constraint satisfaction problem harder. For these
experiments, we use Choco~4.10.2 \citep{choco} with Java~8 on a computer with an
Intel Core i5-8250U processor. For $|\predicates{}|$, $|\variables{}|$,
$|\constants{}|$, $\maxNumNodes{}$, and $\maxNumClauses{} - |\predicates{}|$
(i.e., the number of clauses in addition to the mandatory $|\predicates{}|$
clauses), we assign all combinations of 1, 2, 4, 8. $\maxArity{}$ is assigned to
values 1--4. For each $|\predicates{}|$, we also iterate over all possible
numbers of independent pairs of predicates, ranging from 0 up to
$\binom{|\predicates{}|}{2}$. For each combination of the above-mentioned
parameters, we pick ten random ways to assign arities to predicates (such that
$\maxArity{}$ occurs at least once) and ten random combinations of independent
pairs. We then run the solver with a \SI{60}{\second} timeout.

@@

\section{CONCLUSIONS}
% TODO (Fazl): equations and 'provides no such guarantees' sound more like
% future work / limitation section. Make it sound more positive.

We were able to design an efficient model for generating both logic programs and
probabilistic logic programs. The model avoids unnecessary symmetries, generates
valid programs, and can ensure predicate independence. Note that there is one
kind of symmetries not handled by our model, i.e., logical equivalence. For
example, all three of these formulas are logically equivalent but syntactically
different: $\neg(\mathsf{P}(X) \lor \mathsf{Q}(X))$, $\neg\mathsf{P}(X) \land
\neg\mathsf{Q}(X)$, and $\neg\mathsf{Q}(X) \land \neg\mathsf{P}(X)$. While there
are many situations where one would prefer to treat logically equivalent
formulas as symmetries that should be eliminated, there are also situations
where we do care about enumerating different ways to express the same
probability distribution or knowledge base. For instance, one could look for the
`best' logically equivalent formula, or investigate whether some formulas result
in faster inference than others.

Another issue worth discussing is that of randomness. Most random models sample
from a well-defined probability distribution whereas a constraint model, while
potentially sufficiently random, provides no such guarantees. In future work, it
would be interesting to examine the probability distribution of logic programs
as generated by our model, although it is not obvious how such a
probability distribution could be characterised.

>>>>

How confidently can we claim that an algorithm works well if it is only tested
on a few types of problems? Perhaps the `inferior' algorithm is actually
better in some specific circumstances. Maybe there are identifiable
types of inputs that make every algorithm default to exponential behaviour
that could be easily overcome with the right strategy. At present, most
inference algorithms for probabilistic logic programs are only evaluated on at
most four different problems: sometimes just a single network
\citep{DBLP:journals/tplp/KimmigDRCR11,DBLP:journals/corr/abs-1112-3785},
sometimes two
\citep{DBLP:journals/corr/abs-1009-3798,DBLP:conf/ecai/BruynoogheMKGVJR10} or
four networks \citep{DBLP:conf/ijcai/VlasselaerBKMR15}, all coming from a range
of areas such as social networks and citation/genetic/biological data sets. To
better understand the strengths and weaknesses of these algorithms, along with
real data they should also be evaluated on a range of synthetic problems that
accurately represent the potential complexities that have to be handled by a
well-designed inference algorithm. Furthermore, random logic program generators
can be useful in combination with methods that generate random data with
probabilistic logic programs \citep{DBLP:conf/soict/Dries15} and can be used as
a component of learning.

Most current approaches to generating random logic programs are restrictive,
e.g., limited to clauses with only two literals
\citep{DBLP:conf/lpnmr/NamasivayamT09}, or to clauses of the form $a \gets \neg
b$ \citep{DBLP:journals/tocl/WenWSL16}, but others are more expressive, e.g.,
defining a program only by the (maximum) number of atoms in the body and the
total number of rules \citep{DBLP:conf/iclp/ZhaoL03}. We introduce a way to
generate random logic programs using a constraint model. The same model can
generate both probabilistic programs directly in the syntax of ProbLog
\citep{DBLP:conf/ijcai/RaedtKT07} as well as non-probabilistic Prolog programs.
For generated probabilistic programs to be valid, we use a custom constraint to
detect negative cycles (as described in \cref{sec:cycles}). A major advantage of
our constraint-based approach is that one can easily add additional constraints
to the model. To demonstrate that, in \cref{sec:independence} we present a
custom constraint with propagation and entailment algorithms that can ensure
predicate independence. \cref{sec:counting} also presents a combinatorial
argument for correctness, where we produce combinatorial expressions that count
the number of programs that the model should produce for various parameter
values. Finally, \cref{sec:experiments} shows how the model scales when tasked
with producing more complicated programs and identifies the relationships
between parameter values and the empirical hardness of the program generation
task.

Overall, our main contributions are concerned purely with logic programming-based
languages and frameworks, which capture a major fragment of statistical
relational learning \citep{DBLP:series/synthesis/2016Raedt}.
However, since probabilistic logic programming languages are closely related
to other areas of machine learning, including (imperative) probabilistic
programming \citep{DBLP:journals/ml/RaedtK15}, our results can lay the
foundations for exploring broader questions on generating and testing
programs/algorithms in machine learning.

@@

We need to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate unnecessary symmetries
and adhere to our desired format. First, we can recognise that the order of the
elements in the $\variable{structure}$ array does not matter, i.e., the
structure is only defined by how the elements link to each other, so we can add
a constraint saying that:

@@

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of thrashing,
i.e., situations where a contradiction is being `fixed' by making changes that
have no hope of fixing the contradiction. Thus, we partition all decision
variables into an ordered list of groups, and require the values of all
variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into:
\begin{enumerate*}
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 20, 40,
80, \dots$ contradictions.

We ran close to \num{400000} experiments, investigating whether the model is
efficient enough to generate reasonably-sized programs and gaining insight into
what parameter values make the constraint satisfaction problem harder. For these
experiments, we use Choco~4.10.2 \citep{choco} with Java~8. For
$|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$, and
$\maxNumClauses{} - |\predicates{}|$ (i.e., the number of clauses in addition to
the mandatory $|\predicates{}|$ clauses), we assign all combinations of 1, 2, 4,
8. $\maxArity{}$ is assigned to values 1--4. For each $|\predicates{}|$, we also
iterate over all possible numbers of independent pairs of predicates, ranging
from 0 up to $\binom{|\predicates{}|}{2}$. For each combination of the
above-mentioned parameters, we pick ten random ways to assign arities to
predicates (such that $\maxArity{}$ occurs at least once) and ten random
combinations of independent pairs. We then run the solver with a
\SI{60}{\second} timeout.

@@

\section{CONCLUSIONS \& FUTURE WORK}

We were able to design an efficient model for generating both logic programs and
probabilistic logic programs. The model avoids unnecessary symmetries, generates
valid programs, and can ensure predicate independence. Note that our model
treats logically equivalent but syntactically different formulas as different.
While there are many situations where one would prefer to eliminate them like
symmetries, there are also situations where we do care about enumerating
different ways to express the same probability distribution or knowledge base.
For instance, one could look for the `best' logically equivalent formula, or
investigate whether some formulas result in faster inference than others.

Another topic worth mentioning is randomness. Most random models sample from a
well-defined probability distribution whereas a constraint model, while
sufficiently random, provides no such guarantees. In future work, it would be
interesting to examine the probability distribution of logic programs as
generated by our model, although it is not obvious how such a probability
distribution could be characterised.
